experDesign: stratifying samples into batches with minimal bias

### 2021-07-20 paper.Rmd
Update paper and add figures

---
title: 'experDesign: stratifyuing samples into batches with minimal bias'
output:
  word_document:
    keep_md: yes
  md_document:
    preserve_yaml: yes
  pdf_document: default
  html_document:
    df_print: paged
date: "15th July 2021"
affiliations:
- name: Centro de Investigación Biomédica en Red, Enfermedades Hepáticas y Digestivas
  index: 1
- name: Institut d'Investigacions Biomèdiques August Pi i Sunyer, IDIBAPS
  index: 2
authors:
- name: Lluís Revilla Sancho
  orcid: 0000-0001-9747-2570
  affiliation: 1, 2
- name: Juan-José Lozano
  orcid: 0000-0001-9747-2570
  affiliation: '1'
- name: Azucena Salas
  orcid: 0000-0001-9747-2570
  affiliation: '2'
tags:
- R
- batch effect
- experiment design
bibliography: paper.bib
---

@@

To properly design an experiment, the source of the variation between samples must be identified.
Typically, one can control the environment in which the study or experiment is being conducted.
Sometimes, however, this is not possible, and then one needs to apply certain techniques to control for such variations.
There are three techniques used to decrease the uncertainty about the unwanted variation: blocking, randomization and replication [@klaus2015].

**Blocking** groups samples that are equal according to one or more variables allowing one to estimate the differences between each batch by comparing measurements within blocks.
**Randomization** by mixing the potential confounding variables minimizes the variation in the measurements .
**Replication** increases the number of samples used in an experiment to better estimate the variation of the experiment.
In some settings these techniques can be applied together to ensure the robustness of the study.

Between the designing of an experiment and the measurement of the samples, some samples might be lost, be contaminated, or degradation below quality control threshold can happen.
Even if this does not happen, experiments will occasionally need to be carried out in batches.
Due to technical, for example, the machine cannot measure more than one sample at a time; and practical reasons, for instance, it may not be possible to obtain additional measurements in the field during the allotted time.

@@

To prevent the batch effect to confound your analysis after the initial design of the experiment, there are two options: randomization and replication.
Randomization, can help reduce variations across groups, while replication helps estimate the variation of the measurements or samples, thus increasing the precision of the estimates of the true value on the analysis.
Replications consist of increasing the number of measurements with similar attributes.
When there is one extraction and then a sample is measured multiple times, this is referred to as a technical replicate.
Technical replicates help estimate the variation of the measurement method, and thus the possible batch effect.

Randomization and replicates can be used to prevent batch effects to confound the analysis.
However, by examining how the variables are distributed across each batch sample, proper randomization can be ensured, thus minimizing batch effects.
This is known as randomized block experimental design or stratified random sampling experimental design.

@@

-   *minDiff*, at [github](https://github.com/m-Py/minDiff), and its successor *anticlust*, at [CRAN](https://cran.r-project.org/package=anticlust), divide the samples into similar groups, ensuring similarity by enforcing heterogeneity within groups [@papenberg2020].
    Conceptually it is similar to the clustering methods k-means.

@@

There is a bug on Omixer that prevents it to work unless specific conditions are meet that prevents it comparing with the other tools using the same settings:

@@

In ***experDesign*** the difference is used to compare with the original distribution, this approach is similar to the *anticlust* method used for maximum variance, but instead of the difference with the original distribution the objective function is the sum of the squared errors between cluster centers and individual data points.
But instead of comparing to the clusters center it is compared to the whole data.
If the experiment is carried out with a specific spatial distribution, the `spatial` function also distributes the samples homogeneously by position.
See below for an example.

In addition to distributing the samples into batches, ***experDesign*** provides tools to add technical replicates.
When the design is complete, the replicates are referred to as technical replicates [@blainey2014].
To select the technical replicates needed and in order to choose from which samples they are calculated, the function `extreme_cases` is provided.
For easier usage, the `replicates` function designs an experiment with the number of replicates per batch desired.

>>>>

---
title: 'experDesign: stratifying samples into batches with minimal bias'
output:
  word_document:
    keep_md: yes
  md_document:
    preserve_yaml: yes
  pdf_document: default
  html_document:
    df_print: paged
date: "15th July 2021"
affiliations:
- name: Centro de Investigación Biomédica en Red, Enfermedades Hepáticas y Digestivas
  index: 1
- name: Institut d'Investigacions Biomèdiques August Pi i Sunyer, IDIBAPS
  index: 2
authors:
- name: Lluís Revilla Sancho
  orcid: 0000-0001-9747-2570
  affiliation: 1, 2
- name: Juan-José Lozano
  orcid: 0000-0001-9747-2570
  affiliation: '1'
- name: Azucena Salas
  orcid: 0000-0001-9747-2570
  affiliation: '2'
tags:
- R
- batch effect
- experiment design
bibliography: paper.bib
---

@@

To properly design an experiment, the source of the variation between samples must be identified.
Typically, one can control the environment in which the study or experiment is being conducted.
Sometimes, however, this is not possible, and then one needs to apply certain techniques to control for such variations.
There are three techniques used to decrease the uncertainty of the unwanted variation: blocking, randomization and replication [@klaus2015].

**Blocking** groups samples that are equal according to one or more variables, allowing one to estimate the differences between each batch by comparing measurements within blocks.
**Randomization** minimizes the variation in the measurements by mixing the potential confounding variables.
**Replication** increases the number of samples used in an experiment to better estimate the variation of the experiment.
In some settings these techniques can be applied together to ensure the robustness of the study.

Between the designing of an experiment and the measurement of the samples, some samples might be lost, be contaminated, or degradation below quality threshold can happen.
Even if this does not happen, experiments will occasionally need to be carried out in batches.
Batches might be due to technical reasons, for example, the machine cannot measure more than some amount of samples at a time; and practical reasons, for instance, it may not be possible to obtain additional measurements in the field during the allotted time.

@@

To prevent the batch effect to confound an analysis after the initial design of the experiment, there are two options: randomization and replication.
Randomization, can help reduce variations across groups, while replication helps estimate the variation of the measurements or samples, thus increasing the precision of the estimates of the true value on the analysis.
Replications consist of increasing the number of measurements with similar attributes.
When a sample is measured multiple times, this is referred to as a technical replicate.
Technical replicates help estimate the variation of the measurement method, and thus the possible batch effect [@blainey2014].

Randomization and replicates can be used to prevent batch effects to confound the analysis.
However, by examining how the variables are distributed across each batch, proper randomization can be ensured, thus minimizing batch effects.
This is known as randomized block experimental design or stratified random sampling experimental design.

@@

-   *minDiff*, at [github](https://github.com/m-Py/minDiff), and its successor *anticlust*, at [CRAN](https://cran.r-project.org/package=anticlust), divides the samples into similar groups, ensuring similarity by enforcing heterogeneity within groups [@papenberg2020].
    Conceptually it is similar to the clustering methods k-means.

@@

There is a bug on Omixer that prevents it to work unless specific conditions are meet.
That prevents it comparing with the other tools using the same settings:

@@

In ***experDesign*** the difference is used to compare with the original distribution, this approach is similar to the *anticlust* method used for maximum variance, but instead of the difference with the original distribution the objective function is the sum of the squared errors between cluster centers and individual data points and instead of comparing to the clusters center it is compared to the whole data.
If the experiment is carried out with a specific spatial distribution, the `spatial` function also distributes the samples homogeneously by position.
See below for an example.

In addition to distributing the samples into batches, ***experDesign*** provides tools to add technical replicates.
In order to choose them from the available samples, the function `extreme_cases` is provided.
For easier usage, the `replicates` function designs an experiment with the number of replicates per batch desired.
